Population scale mapping of transposable element diversity reveals links to gene regulation and epigenomic variation

### 2016-09-02 Paper/paper.tex
Updates to text

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and comprise a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (type I
elements), which transpose via an RNA intermediate requiring a reverse
transcription reaction, and DNA transposons (type II elements), which
transpose via either a cut-paste or, in the case of Helitrons, a rolling
circle mechanism with no RNA intermediate \cite{Wicker:2007en}. TE
activity poses mutagenic potential as a TE insertion may disrupt
functional regions of the genome. Consequently, safeguard mechanisms
have evolved to suppress this activity, including the methylation of
cytosine nucleotides (DNA methylation) to produce 5-methylcytosine (mC),
a modification that can induce transcriptional silencing of the
methylated locus. In \emph{Arabidopsis thaliana }(Arabidopsis), DNA
methylation occurs in all three DNA sequence contexts: mCG, mCHG, and
mCHH, where H is any base but G. Establishment of DNA methylation marks
can be carried out by two distinct pathways -- the RNA-directed DNA
methylation pathway guided by 24 nucleotide (nt) small RNAs (smRNAs),
and the DDM1/CMT2 pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major
function of DNA methylation in Arabidopsis is in the transcriptional
silencing of TEs. Loss of DNA methylation due to mutations in genes
essential for its establishment or maintenance leads to expression of
previously silent TEs, and in some cases transposition
\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu, Zemach:2013dj}.

@@

The tools developed previously for identification of novel TE insertion
events have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, are not designed with
population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE variant identification
method to existing genome resequencing data for 216 different
Arabidopsis accessions \cite{Schmitz:2013iu}, identifying widespread TE
variation amongst these accessions and enabling exploration of TE
diversity and links to gene regulation and epigenomic variation.

@@

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the Landsberg
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome, generated by \emph{de novo} assembly using Pacific Biosciences
P5-C3 chemistry with a 20 kb insert library \cite{Chin:2013iw}, using
the same alignment parameters as were used to map reads to the Col-0
reference genome. This resulted in 98.7\% of reads being aligned
concordantly to the L\emph{er} reference, whereas 100\% aligned
discordantly or as split reads to the Col-0 reference genome (Table \ref{table1}).
To find whether reads mapped to homologous regions in both the Col-0 and
L\emph{er} reference genomes, we conducted a blast search
\cite{Camacho:2009fc} using the DNA sequence between read pair mapping
locations in the L\emph{er }genome against the Col-0 genome, and found
the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. Thus, reads providing evidence for TE variants map
discordantly or as split reads when mapped to the Col-0 reference
genome, but map concordantly to homologous regions of the L\emph{er} \emph{de
novo} assembled reference genome, indicating that structural variation
is present at the sites identified by TEPID, and that this is resolved
in the \emph{de novo} assembled genome.

@@

TEPID was used to analyze previously published 100 bp paired-end
genome resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
source data 1) and 8,088 TE absence variants (Figure 2 - source data
2) relative to the Col-0 reference accession, totalling 23,095 unique
TE variants. In most accessions TEPID identified 300-500 TE insertions
(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
majority of which were shared by two or more accessions (Figure 2 -
figure supplement \hyperref[fig2s2]{2}).  PCR validations were
performed for a random subset of 10 insertion and 10 absence variants
in 14 accessions (totalling 280 validations), confirming the high
accuracy of TE variant discovery using the TEPID package, with a false
positive rate for both TE insertion and TE absence identification of
\textasciitilde{}9\%, similar to that observed using simulated data
and the L\emph{er} genome analysis (Figure 2 - figure supplement
\hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement
\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
absence calls is not limited by sequencing depth, while TE insertion
identification benefits from high sequencing depth.  However,
accessions with low coverage gained more TE insertion calls during the
TEPID refinement step (Figure 2 - figure supplement
\hyperref[fig2s4]{4}C), indicating that these false negatives were
effectively reduced by leveraging TE variant information for the whole
population.

@@

TE insertions and deletions were distributed throughout chromosome 1
in a pattern that was similar to the distribution of all Col-0 TEs
(Figure \ref{fig2}A). TE deletions and common TE variants were found
in similar chromosomal regions, as deletion variants represent the
rare loss of common variants. TE deletions and common variants were
more highly enriched in the pericentromeric regions than rare variants
or TE insertions. Among TE deletions, type II elements were slightly
less biased towards the centromeres in comparison to the distribution
of type I elements (Figure 2 - figure supplement
\hyperref[fig2s5]{5}). The distribution of rare TE variants and TE
insertions was similar to that observed for regions of the genome
previously identified as being differentially methylated in all DNA
methylation contexts (mCG, mCHG, mCHH) between the wild accessions
(population C-DMRs), while population CG-DMRs, differentially
methylated in the mCG context, less frequently overlapped with all
types of TE variants identified \cite{Schmitz:2013iu}. Furthermore, TE
variants were depleted within genes and DNase I hypersensitivity sites
\cite{Sullivan:2014ii}, while they were enriched in gene flanking
regions and within other annotated TEs or pseudogenes (Figure
\ref{fig2}B). TE deletions and common TE variants were enriched within
the set of TE variants found in gene bodies (Figure 2C, D). No
significant enrichment was found for TE variants within the \emph{KNOT
  ENGAGED ELEMENT (KEE) }regions, previously identified as regions
that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 - figure
supplement \hyperref[fig2s6]{6}). This may indicate that these regions
do not act as a ``TE sink'' as has been previously proposed, or that
the ``TE sink'' activity is restricted to very recent insertions, as
the insertions we analysed in this population were likely older than
those used in the \emph{KEE} study \cite{Grob:2014kg}.

@@

Although thousands of TE variants were identified, they may be linked
to the previously identified single nucleotide polymorphisms (SNPs),
or unlinked from SNPs across the accessions. We tested how frequently
common TE variants (\textgreater{}3 \% MAF, \textgreater{}7
accessions) were linked to adjacent SNPs to determine when they would
represent a previously unassessed source of genetic variation between
accessions.  SNPs that were previously identified between the
accessions \cite{Schmitz:2013iu} were compared to the presence/absence
of individual TE variants.  For the testable TE variants in the
population, the nearest 300 flanking SNPs upstream and 300 SNPs
downstream of the TE variant site were analyzed for local linkage
disequilibrium (LD, r\textsuperscript{2}; see methods). TE variants
were classified as being either `low', `mid', or `high' LD variants by
comparing ranked r\textsuperscript{2} values of TE variant to SNPs
against the median ranked r\textsuperscript{2} value for all between
SNP comparisons (SNP-SNP) to account for regional variation in the
extent of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
variation or selection \cite{Horton:2012jh}. The majority (61\%) of
testable TE variants had low LD with nearby SNPs, and represent a
source of genetic diversity not previously assessed by SNP-based
genotype calling methods (Figure \ref{fig3}C).  29\% of TE variants
displayed high levels of LD and are tagged by nearby SNPs, while only
10\% had intermediate levels of LD. We observed a positive correlation
between TE variant MAF and LD state, with variants of a high minor
allele frequency more often classified as high-LD (Figure
\ref{fig3}D). While the proportion of TE variants classified as high,
mid, or low-LD was mostly the same for both TE insertions and TE
deletions, TE variants with a high MAF (\textgreater{}20\%) that were
unable to be classified as either true deletions or as new insertions
had a much higher proportion of high-LD variants (Figure \ref{fig3}E).
This was consistent with the observation that the more common alleles
were more often in a high-LD state.  TE variants displayed a similar
distribution over chromosome 1 regardless of linkage classification
(Figure 3 - figure supplement \hyperref[fig3s1]{1}). Overall, this
analysis revealed an abundance of previously uncharacterized genetic
variation that exists amongst Arabidopsis accessions caused by the
presence or absence of TEs, and illustrates the importance of
identifying TE variants alongside other genetic diversity such as
SNPs.

@@

As TEs are frequently highly methylated in Arabidopsis
\cite{Zhang:2006eo, Zilberman:2006hx, Cokus:2008fc, Lister:2008bh},
the DNA methylation state surrounding TE variant sites was assessed to
determine whether TE variants might be responsible for differences in
DNA methylation patterns previously observed between the wild
accessions \cite{Schmitz:2013iu}. TE variants were often physically
close to DMRs (Figure \ref{fig6}A). Furthermore, C-DMRs were more
often close to a TE variant than expected, whereas CG-DMRs were rarely
close to TE insertions or TE deletions (Table \ref{table3}). Overall,
54\% of the 13,482 previously reported population C-DMRs were located
within 1 kb of a TE variant (predominantly TE insertions), while only
15\% of CG-DMRs were within 1 kb of a TE variant (Table
\ref{table3}). For C-DMRs, this was significantly more than expected
by chance, while it was significantly less than expected for CG-DMRs
(p \textless{} $1 \times10^{-4}$, determined by resampling 10,000 times). Of
the C-DMRs that were not close to a TE variant, 3,701 (27\% of all
C-DMRs) were within 1 kb of a non-variable TE. Thus, 81\% of C-DMRs
are within 1 kb of a TE when considering both fixed and variable TEs
in the population. Of the remaining 19\% of C-DMRs, most were found in
genes or intergenic regions.

@@

These results indicate that local DNA methylation patterns are
influenced by the differential TE content between genomes, and that
the DNA methylation-dependent silencing of TEs may frequently lead to
the formation of DMRs between wild Arabidopsis accessions. TE
insertions appear to be important in defining local patterns of DNA
methylation, while DNA methylation levels often remain elevated
following a TE deletion, and so are independent from the presence or
absence of TEs in these cases. Importantly, the distance from a TE
insertion to the centromere appears to have a strong impact on whether
an alteration of local DNA methylation patterns will occur. This is
likely due to flanking sequences being highly methylated in the
pericentromeric regions, and so the insertion of a TE cannot further
increase levels of DNA methylation. Overall, a large fraction of the
population C-DMRs previously identified between wild accessions are
correlated with the presence of local TE variants. CG-DMR methylation
levels appear to be mostly independent from the presence/absence of
common TE variants, while rare TE variants have an impact on DNA
methylation levels at both C-DMRs and CG-DMRs. This analysis was aided
by the high sensitivity of TEPID to detect TE variants, enabling the
identification of 8x more variants than has been reported previously
in the population \cite{Quadrana:2016bi}, allowing a thorough
assessment of the impact of TE variants on DNA methylation patterns.

@@

To further investigate the effects of TE variants upon local and
distant DNA methylation levels in the genome, an association scan was
conducted for all common TE variants (\textgreater{}3\% MAF) and all
population C-DMRs for the 124 accessions with both DNA methylation and
TE variant data available. To test the significance of each pairwise
correlation, bootstrap p-value estimates were collected based on 500
permutations of accession labels. TE-DMR associations were deemed
significant if they had an association more extreme than any of the
500 permutations (p \textless{} 1/500). A band of significant
associations was observed for TE insertions and their nearby C-DMRs,
signifying a local association between TE insertion presence/absence
and C-DMR methylation (Figure \ref{fig8}A).  This local association was
not as strong for TE deletions (Figure \ref{fig8}B), consistent with our above
findings. While TE variants and DNA methylation showed a local
association, it is also possible that TE variation may influence DNA
methylation states more broadly in the genome, perhaps through
production of \emph{trans-}acting smRNAs or inactivation of genes
involved in DNA methylation establishment or maintenance. To identify
any potential enrichment of C-DMRs regulated in \emph{trans}, the
total number of significant associations was summed for each TE
variant across the whole genome (Figure \ref{fig8}A and B, top panels). At
many sites, far more significant associations were found than expected
due to the false positive rate alone. This suggested the existence of
many putative \emph{trans }associations between TE variants and
genome-wide C-DMR methylation levels. These C-DMRs that appeared to be
associated with a TE insertion in \emph{trans} were further
examined\emph{, }checking for TE insertions near these C-DMRs that
were present in the same accessions as the \emph{trans }associated TE,
as these could lead to a false \emph{trans }association. These were
extremely rare, with only 4 such cases for TE insertions, and 38 cases
for TE deletions, and so were unable to explain the high degree of
\emph{trans }associations found. Overall, this analysis suggests that
certain TE variants may affect DNA methylation levels more broadly in
the genome, as their effects upon DNA methylation are not necessarily
limited to nearby DNA sequences.

@@

Here we have discovered widespread differential TE content between wild
Arabidopsis accessions, and explored the impact of these variants upon
transcription and DNA methylation at the level of individual accessions.
Most TE variants were due to the \emph{de novo }insertion of TEs, while
a smaller subset was likely due to the deletion of ancestral TE copies,
mostly around the pericentromeric regions. A subset (32\%) of TE
variants with a minor allele frequency above 3\% were able to be tested
for linkage with nearby SNPs. The majority of these TE variants were not
in LD with SNPs, indicating that they represent genetic variants
currently overlooked in genomic studies. A marked depletion of TE
variants within gene bodies and DNase I hypersensitivity sites (putative
regulatory regions) is consistent with the more deleterious TE
insertions being removed from the population through selection. Of those
TE variants found in gene bodies, TE deletions were overrepresented,
indicating that the loss of ancestral TEs inserted within genes may be
more frequent, or perhaps less deleterious, than the \emph{de novo
}insertion of TEs into genes.

The high sensitivity of our TE variant detection method allowed
identification of 8x more variants than has been reported previously
\cite{Quadrana:2016bi}, and the identification of such a large number of
TE variants (23,095) in this population gave an opportunity to form
statistically robust correlations between TE presence/absence and
transcript abundance from nearby genes, as well as genome-wide patterns
of DNA methylation. Examples were identified where TE variants appear to
have an effect upon gene expression, both in the disruption of
transcription and in the spreading or disruption of regulatory
information leading to the transcriptional activation of genes,
indicating that these TE variants can have important consequences upon
the expression of protein coding genes (Figure \ref{fig5}). In one case, these
changes in gene expression could be linked with phenotypic changes, with
accessions containing a TE insertion more frequently sensitive to
bacterial infection. Further experiments will be needed to establish a
causal link between this TE insertion and the associated phenotype. An
analysis of rare TE variants, present at a low MAF, further strengthened
this relationship between TE presence/absence and altered transcript
abundance, as a strong enrichment of rare TE variants in accessions with
extreme gene expression ranks in the population was identified.

Importantly, we provide evidence that differential TE content between
genomes of Arabidopsis accessions underlies a large fraction of the
previously reported population C-DMRs. Thus, the frequency of pure
epialleles, independent of underlying genetic variation, may be even
more rare than previously anticipated \cite{Richards:2006br}. Overall,
81\% of all C-DMRs were within 1 kb of a TE, when considering both fixed
and variable TEs in the population, a much higher percentage than has
been reported previously \cite{Schmitz:2013iu, Quadrana:2016bi}. We did
not find evidence of CG-DMR methylation level being altered by the
presence of common TE variants, although rare TE variants may be more
important in shaping patterns of DNA methylation at some CG-DMRs, though
the reasons for this distinction remain unclear. The level of local DNA
methylation changes associated with TE variants was also related to the
distance from a TE variant to the centromere, with variants in the
chromosome arms being more strongly correlated with DNA methylation
levels. This seems to be due to a higher baseline level of DNA
methylation at the pericentromeric regions, which prevent any further
increase in DNA methylation level following insertion of a TE.
Furthermore, we found an important distinction between TE insertions and
TE deletions in the effect that these variants have on nearby DNA
methylation levels. While flanking DNA methylation levels increase
following a TE insertion, the deletion of an ancestral TE was often not
associated with a corresponding decrease in flanking DNA methylation
levels (Figure \ref{fig7}). This indicates that high levels of DNA methylation,
once established, may be maintained in the absence of the TE insertion
that presumably triggered the original change in DNA methylation level.
It is then possible that TE variants explain more of the inter-accession
variation in DNA methylation patterns than we find direct evidence for,
if some C-DMRs were formed by the insertion of an ancestral TE that is
now absent in all the accessions analysed here. These DMRs would then
represent the epigenetic ``scars'' of past TE insertions.

>>>>

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and comprise a large fraction of most
eukaryotic genomes. The two types of TEs are retrotransposons (type I
elements), which transpose via an RNA intermediate requiring a reverse
transcription reaction, and DNA transposons (type II elements), which
transpose via either a cut-paste or, in the case of Helitrons, a
rolling circle mechanism with no RNA intermediate
\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
insertion may disrupt functional regions of the genome. Consequently,
safeguard mechanisms have evolved to suppress this activity, including
the methylation of cytosine nucleotides (DNA methylation) to produce
5-methylcytosine (mC), a modification that can induce transcriptional
silencing of the methylated locus. In \emph{Arabidopsis thaliana}
(Arabidopsis), DNA methylation occurs in all three DNA sequence
contexts: mCG, mCHG, and mCHH, where H is any base but
G. Establishment of DNA methylation marks can be carried out by two
distinct pathways -- the RNA-directed DNA methylation pathway guided
by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
methylation in Arabidopsis is in the transcriptional silencing of
TEs. Loss of DNA methylation due to mutations in genes essential for
its establishment or maintenance leads to expression of previously
silent TEs, and in some cases transposition \cite{Mirouze:2009km,
  Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,
  Zemach:2013dj}.

@@

The tools developed previously for identification of novel TE
insertion events have several limitations. They either require a
library of active TE sequences, cannot identify TE absence variants,
are not designed with population studies in mind, or specifically aim
to reduce false-positives rather than false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel
algorithm, TEPID (Transposable Element Polymorphism IDentification),
which is designed for population studies. We tested our algorithm
using both simulated and real Arabidopsis sequencing data, finding
that TEPID is able to accurately identify TE presence/absence variants
with respect to the Col-0 reference genome. We applied our TE variant
identification method to existing genome resequencing data for 216
different Arabidopsis accessions \cite{Schmitz:2013iu}, identifying
widespread TE variation amongst these accessions and enabling
exploration of TE diversity and links to gene regulation and
epigenomic variation.

@@

In order to further assess the sensitivity and specificity of TE
variant discovery using TEPID, we identified TE variants in the
Landsberg \emph{erecta} (L\emph{er}) accession, and compared these
with the L\emph{er }genome assembly created using long PacBio
sequencing reads \cite{Chin:2013iw}. Previously published 100 bp
paired-end L\emph{er} genome resequencing reads
\cite{Schneeberger:2011ft} were first analyzed using TEPID, enabling
identification of 446 TE insertions (Figure 1 - source data 1) and 758
TE absence variants (Figure 1 - source data 2) with respect to the
Col-0 reference TE annotation. Reads providing evidence for these
variants were then mapped to the L\emph{er} reference genome,
generated by \emph{de novo} assembly using Pacific Biosciences P5-C3
chemistry with a 20 kb insert library \cite{Chin:2013iw}, using the
same alignment parameters as were used to map reads to the Col-0
reference genome. This resulted in 98.7\% of reads being aligned
concordantly to the L\emph{er} reference, whereas 100\% aligned
discordantly or as split reads to the Col-0 reference genome (Table
\ref{table1}). To find whether reads mapped to homologous regions in both the
Col-0 and L\emph{er} reference genomes, we conducted a blast search
\cite{Camacho:2009fc} using the DNA sequence between read pair mapping
locations in the L\emph{er} genome against the Col-0 genome, and found
the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence
variants in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. Thus, reads providing evidence for TE variants map
discordantly or as split reads when mapped to the Col-0 reference
genome, but map concordantly to homologous regions of the L\emph{er}
\emph{de novo} assembled reference genome, indicating that structural
variation is present at the sites identified by TEPID, and that this
is resolved in the \emph{de novo} assembled genome.

@@

TEPID was used to analyze previously published 100 bp paired-end
genome resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
source data 1) and 8,088 TE absence variants (Figure 2 - source data
2) relative to the Col-0 reference accession, totalling 23,095 unique
TE variants.  A recent study focused on identifying recent TE
insertions containing target site duplications in this population
\cite{Quadrana:2016bi}.  Our goal was to provide a comprehensive
assessment of TE presence/absence variation in Arabidopsis.  In most
accessions TEPID identified 300-500 TE insertions (mean = 378) and
1,000-1,500 TE absence variants (mean = 1,279), the majority of which
were shared by two or more accessions (Figure 2 - figure supplement
\hyperref[fig2s2]{2}).  PCR validations were performed for a random
subset of 10 insertion and 10 absence variants in 14 accessions
(totalling 280 validations), confirming the high accuracy of TE
variant discovery using the TEPID package, with a false positive rate
for both TE insertion and TE absence identification of
\textasciitilde{}9\%, similar to that observed using simulated data
and the Ler genome analysis (Figure 2 - figure supplement
\hyperref[fig2s3]{3}).  The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement
\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
absence calls is not limited by sequencing depth, while TE insertion
identification benefits from high sequencing depth. However,
accessions with low coverage gained more TE insertion calls during the
TEPID refinement step (Figure 2 - figure supplement
\hyperref[fig2s4]{4}C), indicating that these false negatives were
effectively reduced by leveraging TE variant information for the whole
population.

@@

TE insertions and deletions were distributed throughout chromosome 1
in a pattern that was similar to the distribution of all Col-0 TEs
(Figure \ref{fig2}A). TE deletions and common TE variants were found
in similar chromosomal regions, as deletion variants represent the
rare loss of common variants.  TE deletions and common variants were
more highly enriched in the pericentromeric regions than rare variants
or TE insertions.  Among TE deletions, type II elements were slightly
less biased towards the centromeres in comparison to the distribution
of type I elements (Figure 2 - figure supplement
\hyperref[fig2s5]{5}).  The distribution of rare TE variants and TE
insertions was similar to that observed for regions of the genome
previously identified as being differentially methylated in all DNA
methylation contexts (mCG, mCHG, mCHH) between the wild accessions
(population C-DMRs), while population CG-DMRs, differentially
methylated in the mCG context, less frequently overlapped with all
types of TE variants identified \cite{Schmitz:2013iu}. Furthermore, TE
variants were depleted within genes and DNase I hypersensitivity sites
\cite{Sullivan:2014ii}, while they were enriched in gene flanking
regions and within other annotated TEs or pseudogenes (Figure
\ref{fig2}B). TE deletions and common TE variants were enriched within
the set of TE variants found in gene bodies (Figure \ref{fig2}C,
D). No significant enrichment was found for TE variants within the
\emph{KNOT ENGAGED ELEMENT (KEE) }regions, previously identified as
regions that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 -
figure supplement \hyperref[fig2s6]{6}). This may indicate that these
regions do not act as a ``TE sink'' as has been previously proposed,
or that the ``TE sink'' activity is restricted to very recent
insertions, as the insertions we analysed in this population were
likely older than those used in the \emph{KEE} study
\cite{Grob:2014kg}.

@@

Although thousands of TE variants were identified, they may be linked
to the previously identified single nucleotide polymorphisms (SNPs),
or unlinked from SNPs across the accessions. We tested how frequently
common TE variants (\textgreater{}3\% MAF, \textgreater{}7 accessions)
were linked to adjacent SNPs to determine when they would represent a
previously unassessed source of genetic variation between accessions.
SNPs that were previously identified between the accessions
\cite{Schmitz:2013iu} were compared to the presence/absence of
individual TE variants. For the testable TE variants in the
population, the nearest flanking 300 SNPs upstream and 300 SNPs
downstream of the TE variant site were analyzed for local linkage
disequilibrium (LD, r\textsuperscript{2}; see methods). TE variants
were classified as being either `low', `mid', or `high' LD variants by
comparing ranked r\textsuperscript{2} values of TE variant to SNPs
against the median ranked r\textsuperscript{2} value for all between SNP comparisons
(SNP-SNP) to account for regional variation in the extent of SNP-SNP
LD (Figure \ref{fig3}A, B) due to recombination rate variation or
selection \cite{Horton:2012jh}. The majority (61\%) of testable TE
variants had low LD with nearby SNPs, and represent a source of
genetic diversity not previously assessed by SNP-based genotype
calling methods (Figure \ref{fig3}C).  29\% of TE variants displayed
high levels of LD and are tagged by nearby SNPs, while only 10\% had
intermediate levels of LD. We observed a positive correlation between
TE variant MAF and LD state, with variants of a high minor allele
frequency more often classified as high-LD (Figure \ref{fig3}D). While
the proportion of TE variants classified as high, mid, or low-LD was
mostly the same for both TE insertions and TE deletions, TE variants
with a high MAF (\textgreater{}20\%) that were unable to be classified
as either true deletions or as new insertions had a much higher
proportion of high-LD variants (Figure \ref{fig3}E). This was
consistent with the observation that the more common alleles were more
often in a high-LD state. TE variants displayed a similar distribution
over chromosome 1 regardless of linkage classification (Figure 3 -
figure supplement \hyperref[fig3s1]{1}). Overall, this analysis
revealed an abundance of previously uncharacterized genetic variation
that exists amongst Arabidopsis accessions caused by the presence or
absence of TEs, and illustrates the importance of identifying TE
variants alongside other genetic diversity such as SNPs.

@@

As TEs are frequently highly methylated in Arabidopsis
\cite{Zhang:2006eo, Zilberman:2007hx, Cokus:2008fc, Lister:2008bh},
the DNA methylation state surrounding TE variant sites was assessed to
determine whether TE variants might be responsible for differences in
DNA methylation patterns previously observed between the wild
accessions \cite{Schmitz:2013iu}. TE variants were often physically
close to DMRs (Figure \ref{fig6}A). Furthermore, C-DMRs were more
often close to a TE variant than expected, whereas CG-DMRs were rarely
close to TE insertions or TE deletions (Table \ref{table3}). Overall,
54\% of the 13,482 previously reported population C-DMRs were located
within 1 kb of a TE variant (predominantly TE insertions), while only
15\% of CG-DMRs were within 1 kb of a TE variant (Table
\ref{table3}). For C-DMRs, this was significantly more than expected
by chance, while it was significantly less than expected for CG-DMRs
(p \textless{} $1 \times10^{-4}$, determined by resampling 10,000
times). Of the C-DMRs that were not close to a TE variant, 3,701 (27\%
of all C-DMRs) were within 1 kb of a non-variable TE. Thus, 81\% of
C-DMRs are within 1 kb of a TE when considering both fixed and
variable TEs in the population. Of the remaining 19\% of C-DMRs, most
were found in genes or intergenic regions.

@@

These results indicate that local DNA methylation patterns are
influenced by the differential TE content between genomes, and that the
DNA methylation-dependent silencing of TEs may frequently lead to the
formation of DMRs between wild Arabidopsis accessions. TE insertions
appear to be important in defining local patterns of DNA methylation,
while DNA methylation levels often remain elevated following a TE
deletion, and so are independent from the presence or absence of TEs in
these cases. Importantly, the distance from a TE insertion to the
centromere appears to have a strong impact on whether an alteration of
local DNA methylation patterns will occur. This is likely due to
flanking sequences being highly methylated in the pericentromeric
regions, and so the insertion of a TE cannot further increase levels of
DNA methylation. Overall, a large fraction of the population C-DMRs
previously identified between wild accessions are correlated with the
presence of local TE variants. CG-DMR methylation levels appear to be
mostly independent from the presence/absence of common TE variants,
while rare TE variants have an impact on DNA methylation levels at both
C-DMRs and CG-DMRs. This analysis was aided by the high sensitivity of
TEPID to detect TE variants, allowing a thorough assessment of the
impact of TE variants on DNA methylation patterns.

@@

To further investigate the effects of TE variants upon local and
distant DNA methylation levels in the genome, an association scan was
conducted for all common TE variants (\textgreater{}3\% MAF) and all
population C-DMRs for the 124 accessions with both DNA methylation and
TE variant data available. To test the significance of each pairwise
correlation, bootstrap p-value estimates were collected based on 500
permutations of accession labels. TE-DMR associations were deemed
significant if they had an association more extreme than any of the
500 permutations (p \textless{} 1/500). A band of significant
associations was observed for TE insertions and their nearby C-DMRs,
signifying a local association between TE insertion presence/absence
and C-DMR methylation (Figure \ref{fig8}A).  This local association
was not as strong for TE deletions (Figure \ref{fig8}B), consistent
with our above findings. While TE variants and DNA methylation showed
a local association, it is also possible that TE variation may
influence DNA methylation states more broadly in the genome, perhaps
through production of \emph{trans-}acting smRNAs or inactivation of
genes involved in DNA methylation establishment or maintenance. To
identify any potential enrichment of C-DMRs regulated in \emph{trans},
the total number of significant associations was summed for each TE
variant across the whole genome (Figure \ref{fig8}A and B, top
panels). At many sites, far more significant associations were found
than expected due to the false positive rate alone. This suggested the
existence of many putative \emph{trans }associations between TE
variants and genome-wide C-DMR methylation levels. These C-DMRs that
appeared to be associated with a TE insertion in \emph{trans} were
further examined, checking for TE insertions near these C-DMRs that
were present in the same accessions as the \emph{trans }associated TE,
as these could lead to a false \emph{trans }association. These were
extremely rare, with only 4 such cases for TE insertions, and 38 cases
for TE deletions, and so were unable to explain the high degree of
\emph{trans }associations found. Overall, this analysis suggests that
certain TE variants may affect DNA methylation levels more broadly in
the genome, as their effects upon DNA methylation are not necessarily
limited to nearby DNA sequences.

@@

Here we have discovered widespread differential TE content between wild
Arabidopsis accessions, and explored the impact of these variants upon
transcription and DNA methylation at the level of individual accessions.
Most TE variants were due to the \emph{de novo }insertion of TEs, while
a smaller subset was likely due to the deletion of ancestral TE copies,
mostly around the pericentromeric regions. A subset (32\%) of TE
variants with a minor allele frequency above 3\% were able to be tested
for linkage with nearby SNPs. The majority of these TE variants
exhibited only low levels of LD with nearby SNPs, indicating that they
represent genetic variants currently overlooked in genomic studies. A
marked depletion of TE variants within gene bodies and DNase I
hypersensitivity sites (putative regulatory regions) is consistent with
the more deleterious TE insertions being removed from the population
through selection. Of those TE variants found in gene bodies, TE
deletions were overrepresented, indicating that the loss of ancestral
TEs inserted within genes may be more frequent, or perhaps less
deleterious, than the \emph{de novo }insertion of TEs into genes.

A previous study focused on recent TE insertions in the Arabidopsis
population \cite{Quadrana:2016bi}, thus the extensive variation
between accessions due to older TE insertions or TE deletions has not
been explored. We examined both old and new TE variants (common and
rare alleles), as well as TE deletions, allowing the detection of
23,095 TE variants in the population. This large sample size gave an
opportunity to form statistically robust correlations between TE
presence/absence and transcript abundance from nearby genes, as well
as genome-wide patterns of DNA methylation. Examples were identified
where TE variants appear to have an effect upon gene expression, both
in the disruption of transcription and in the spreading or disruption
of regulatory information leading to the transcriptional activation of
genes, indicating that these TE variants can have important
consequences upon the expression of protein coding genes (Figure
\ref{fig5}). In one case, these changes in gene expression could be
linked with phenotypic changes, with accessions containing a TE
insertion more frequently sensitive to bacterial infection. Further
experiments will be needed to establish a causal link between this TE
insertion and the associated phenotype. An analysis of rare TE
variants, present at a low MAF, further strengthened this relationship
between TE presence/absence and altered transcript abundance, as a
strong enrichment of rare TE variants in accessions with extreme gene
expression ranks in the population was identified.  Therefore, the
effects of TE insertions appear to be long-lasting, as there was
little difference between common (old) and rare (young) variants in
the impact upon gene expression (Figure \ref{fig4}).

Importantly, we provide evidence that differential TE content between
genomes of Arabidopsis accessions underlies a large fraction of the
previously reported population C-DMRs. Thus, the frequency of pure
epialleles, independent of underlying genetic variation, may be even
more rare than previously anticipated \cite{Richards:2006br}. Overall,
81\% of all C-DMRs were within 1 kb of a TE, when considering both fixed
and variable TEs in the population. We did not find evidence of CG-DMR
methylation level being altered by the presence of common TE variants,
although rare TE variants may be more important in shaping patterns of
DNA methylation at some CG-DMRs, though the reasons for this distinction
remain unclear. The level of local DNA methylation changes associated
with TE variants was also related to the distance from a TE variant to
the centromere, with variants in the chromosome arms being more strongly
correlated with DNA methylation levels. This seems to be due to a higher
baseline level of DNA methylation at the pericentromeric regions, which
prevent any further increase in DNA methylation level following
insertion of a TE. Furthermore, we found an important distinction
between TE insertions and TE deletions in the effect that these variants
have on nearby DNA methylation levels. While flanking DNA methylation
levels increase following a TE insertion, the deletion of an ancestral
TE was often not associated with a corresponding decrease in flanking
DNA methylation levels (Figure \ref{fig7}). This indicates that high levels of
DNA methylation, once established, may be maintained in the absence of
the TE insertion that presumably triggered the original change in DNA
methylation level. It is then possible that TE variants explain more of
the inter-accession variation in DNA methylation patterns than we find
direct evidence for, if some C-DMRs were formed by the insertion of an
ancestral TE that is now absent in all the accessions analysed here.
These DMRs would then represent the epigenetic ``scars'' of past TE
insertions.
